---
title: "Imputation methods in mice"
author: "Aaron Smith"
date: "January 6, 2021"
output: html_document
---

The mice package in R has a lot of methods for imputing missing values. Here we look at how some of these methods perform.

# Methods in mice

Method parameter|Data type|Description
-|-|-
2lonly.pmm|any|Level-2 class predictive mean matching
cart|any|Classification and regression trees
midastouch|any|Weighted predictive mean matching
pmm|any|Predictive mean matching
rf|any|Random forest imputations
sample|any|Random sample from observed values
2l.lmer|numeric|Level-1 normal homoscedastic, lmer
2l.norm|numeric|Level-1 normal heteroscedastic
2l.pan|numeric|Level-1 normal homoscedastic, pan
2lonly.mean|numeric|Level-2 class mean
2lonly.norm|numeric|Level-2 class normal
mean|numeric|Unconditional mean imputation
norm|numeric|Bayesian linear regression
norm.boot|numeric|Linear regression using bootstrap
norm.nob|numeric|Linear regression ignoring model error
norm.predict|numeric|Linear regression, predicted values
quadratic|numeric|Imputation of quadratic terms
ri|numeric|Random indicator for nonignorable data
polr|ordered|Proportional odds model
lda|unordered|Linear discriminant analysis
polyreg|unordered|Polytomous logistic regression
2l.bin|binary|Level-1 logistic, glmer
logreg|binary|Logistic regression
logreg.boot|binary|Logistic regression with bootstrap

# Load data

```{r}
#library(printr)
options(digits = 2)
nashville <- as.data.frame(readr::read_csv(
  file = "C:\\Users\\smithaz12\\Desktop\\UCF\\STA6714Spring2021\\input\\Nashville_housing_data_2013_2016.csv",
  col_types = "nncccccDncccccccncnccnnncnccnnn"
))
colnames(nashville) <- make.names(
  names = colnames(nashville)
)
colnames(nashville) <- gsub(
  x = colnames(nashville),
  pattern = "[.]",
  replacement = " "
)
colnames(nashville) <- trimws(gsub(
  x = colnames(nashville),
  pattern = "\\s+",
  replacement = " "
))
colnames(nashville) <- make.names(
  names = colnames(nashville)
)
nashville <- nashville[,!colnames(nashville) %in% c("X1","Unnamed.0","State")]
nashville <- nashville[,order(sapply(
  X = nashville,
  FUN = class
))]
nashville <- nashville[order(apply(
  X = nashville,
  MARGIN = 1,
  FUN = function(x) sum(is.na(x))
)),]
nashville <- nashville[,order(sapply(
  X = nashville,
  FUN = function(x) sum(is.na(x))
))]
rownames(nashville) <- 1:nrow(nashville)
```

# Filter columns to complete columns and numeric columns with missing values

```{r}
v_columns <- c(
  "Sale.Price",
  "Land.Use","Sold.As.Vacant","Multiple.Parcels.Involved.in.Sale",
  "Acreage","Neighborhood","Building.Value","Total.Value",
  "Full.Bath","Finished.Area","Year.Built","Bedrooms","Half.Bath"
)
M <- nashville[,v_columns]
```

# Create initial imputation as starting point, create prediction matrix

```{r}
require(mice)

M_init <- M[,colnames(M) != "Sale.Price"]
for(j in colnames(M_init)){
  if(class(M_init[,j]) == "numeric") M_init[is.na(M_init[,j]),j] <- mean(
    x = M_init[,j],
    na.rm = TRUE
  )
  if(class(M_init[,j]) == "character"){
    M_init[is.na(M_init[,j]),j] <- names(sort(
      x = table(M_init[,j]),
      decreasing = TRUE
    ))[1]
  }
}

mice_mean <- mice::mice(
  data = M[,colnames(M) != "Sale.Price"],
  method = "mean",
  m = 1,
  maxit = 1,
  data.init = M_init,
  seed = 823
)
```

# Impute missing values with mice

```{r}
v_method <- c(
  "pmm","sample","cart","rf",
  "2l.bin",
  "mean","norm","norm.nob","norm.boot","norm.predict","ri","2l.lmer"
)

gc()
list_mice <- lapply(
  X = v_method,
  FUN = function(x) try({
    gc()
    mice::mice(
      data = M[,colnames(M) != "Sale.Price"],
      method = x,
      predictorMatrix = mice_mean$predictorMatrix,
      m = 1,
      maxit = 1,
      data.init = M_init,
      seed = 823
    )
  })
)
gc()
names(list_mice) <- v_method
```

# Extract complete data, append target variable

```{r}
list_complete <- lapply(
  X = list_mice,
  FUN = mice::complete
)
for(j in names(list_complete)) list_complete[[j]] <- data.frame(
  Sale.Price = M$Sale.Price,
  list_complete[[j]]
)
```

# Evaluate imputation using linear regression

```{r}
list_lm <- lapply(
  X = list_complete,
  FUN = function(x) lm(
    data = x,
    formula = log(Sale.Price) ~.
  )
)
v_mae <- sapply(
  X = list_lm,
  FUN = function(x) mean(abs(predict(x) - log(M$Sale.Price)))
)
sort(v_mae)
```